# (PART\*) Part I. R Programming Basics {-}

# R Basics

## Operators  

Let us start with the basic. Almost all programming languages can be used as a calculator. You can use the `+`, `-`, `*` and `/` symbols to perform addition, subtraction, multiplication and division between two numeric values. These are called _arithmetic operators_. Type the following lines of code on the RStudio console pane and check the results of these operations:
```r
2 + 2
10 - 5
4.5*3
4/2
```
The `[1]` in front of the results indicates that the value after it is the first value of the result for that operation.

Some other arithmetic operators are `^` or `**` for exponentiation, `%%` [modulus](https://www.khanacademy.org/computing/computer-science/cryptography/modarithmetic/a/what-is-modular-arithmetic) and `%/%` for [integer division](https://mathworld.wolfram.com/IntegerDivision.html). 

_Logical operators_ perform comparison between two values. Logical operations return either `TRUE` or `FALSE`. Type and run the following lines of code and check the results:
```r
5 < 4 # less than
2 > 1 # greater than
2==1  # exactly equal to
5>=5  # greater than and equal to
5!=6  # not equal to
```

A very useful __binary operator__ in R is denoted by the colon `:`. Binary operators are operators that perform actions with two values. The colon operator generates regular sequences. It can also create sequence in reverse order. Whether the sequence is increasing or decreasing depends on whether the value before the operator is less than or greater than the value after the operator. 
```r
1:15
15:1
```

There are many other operators in R and we learn about them as we go through the remaining topics. 

## Objects  

### Creating Objects  
The power of programming lies in its ability to not only manipulate numeric values but also textual values. Any value can be stored as an object in a computer program and then referenced later for further manipulation. The _assignment operator_, `<-`, is used to assign a value to an object. It comprises the less than symbol `<` and a hyphen `-`. Manipulation to the value can be done by manipulating the object. 
```r
a <- 5
a+5
a*2.5
a>=5
```
Object provide a way of storing data with a descriptive name such that the programs can be easily read and understood by its readers. For example, we can create an object `aapl`, which is the ticker for the company Apple Inc. and assign it a value of `274.5b`, which is Apple's revenue in billions for the year 2020. Now, every time you type `aapl` on the console, it will print `274.5b`, unless you reassign a different value to this object. 
```r
aapl <- "274.5b"
aapl
```

Note: `=` can also be used as an assignment operator but is not recommended for R. In some cases it is forbidden. 

### Naming Objects  
Naming an object is a not that easy. You want the name to be meaningful and easily understood but not very lengthy. Objects with multiple words can be created by joining them with an underscore or a dot. You can also capitalize the different words. Object names cannot start with a number or a dot. Avoid using alphanumeric character in object names. 

Most importantly do not use a name that has already been assigned to some other value, unless you have good reason to do so. For example, variable `pi` exists in R by default. So does keyword variables like `TRUE`, `FALSE`, `NA`, etc. Do not use these names while creating an object. Doing so will override their default value and you will have to restart the R session to revert them back. 
```r
pi
TRUE
FALSE
NA
```
R is case sensitive meaning `TRUE` and `True` are not the same. Try it yourself and see what you get!

So far, you have created two objects, `a` and `aapl`. In the top right hand pane of RStudio, you will see these two objects listed on the 'Environment' tab. If you click on the down arrow by the `List` icon and select `Grid` view, you will see additional details about these objects such as type, length, size and value. 

**Exercise**

Create an object "company" and assign it a value of "AAPL". Create another object "revenue" and assign it a value of 274.5.

```{r}
# your code here
```

## Data Types

Lets look at the type in detail because it is very important to know how R deals with different types of data. Type of an object, also known as its data type, tells us the kind of operations that is possible with that object. For example, `a` has numeric data type and therefore arithmetic operations can be performed to it. `aapl` is of character type and thus arithmetic operators cannot be applied to it. If you do, you will receive an error: 
```r
aapl+1
aapl*3
aapl/3
```
The error message tells you that you are applying a binary operator to a non-numeric value. 

Four types of basic data types we will focus on are - numeric, integer, logical, and character. 

### Numeric  
Numeric data are contains decimal numbers. Although, it may also include values without decimal numbers. In most cases, numeric data is chosen by default in R for numeric values. 

**Exercise**

Check whether the data type of the variable "revenue", which you created in section 1.2, is numeric.

```{r}
# your code here
```

### Integer
Integer are whole numbers, meaning numbers without decimal. 

R provides numerous functions to check the data type of an object and to modify it as well.
```r
is.numeric(a)
is.integer(a)
```
Notice the results of the two operations. Even though `a` does not have any decimal point, R has stored it as numeric instead of integer type. While R is able to automatically distinguish data types in most cases, often you will have to explicitly tell R which data type you want for a given object. 
```r
a <- as.integer(a)
is.integer(a)
```
**Exercise**

Convert the data type of variable "revenue", which you created in Section 1.2, into integer. What is a possible danger of such data conversion?

```{r}
# your code here
```


### Logical  
Logical data take either `TRUE` or `FALSE` values. Logical data type result from logical operations:
```r
logical_expression <- (a > 5)
is.logical(logical_expression)
```
`NA` is also considered to be of logical data type and signifies missing or null values. It is a special type of logical. 
```r
is.logical(NA)
```
**Exercise**

Create a logical expression and check the data type.

```{r}
# your code here
```

### Character 
Finally, character data types represent string. String in computer program refers to any letter, digit, punctuation and extra characters enclosed inside quotation marks. `aapl`, is stored as character because while creating this object we enclosed its values in quotation marks. 
```r
is.character(aapl)
```
Data type modifications must be done with careful consideration to avoid unexpected results. For example, note the data loss occurred as a result of the following data type conversions below:
```r
as.integer(5.9)
as.numeric(aapl)
```

**Exercise**

Check if the data type of variable "company", which you created in Section 1.2 is a character.

```{r}
# your code here
```


## Data Structures  
Now that you have been introduced to objects and data types, let's look at some of the main data structures provided by R to store real world data. 

### Vectors  
Vectors are the simplest data structure in R. Vectors are 1-dimensional data structures. So far the objects we have created are __scalar__ vectors, scalar because they contain a single value. Vectors with more than one value can be created as follows:
```r
is.vector(a)

revenue <- c(386.1, 286.2, 274.5, 182.5, 143, 119.5, 86, 82.6, 31.5, 21.9)
tickers <- c('AMZN', 'BRK.B', 'AAPL', 'GOOG', 'MSFT', 'JPM', 'FB', 'JNJ', 'TSLA', 'V')

print(revenue)
print(tickers)

is.vector(revenue)
is.vector(tickers)
```
The function `length` can be used to get the number of items in a vector.
```r
length(revenue)
length(tickers)
```

Vectors can contain any data type but it must all be of the same type. In other words, a vector cannot have mixtures of data types. If you tried to create a vector with multiple data types, R will coerce the data type to the most flexible data type. There is one exception, which is with the `NA` data type. `NA` can be included within vectors of any other data type to denote missing data in that vector.
```r
mixed_vector <- c(386.1, '286.2', 274.5, NA, 143, 119.5, NA, 82.6, '31.5', '21.9')
typeof(mixed_vector)
```

If you want to __extract only one item__ from the vector, you can do so by typing the name of the vector and then indicating the position of that item in the vector inside square brackets. For example, to extract the first item in the vector "revenue" and "tickers", you can write the following code:
```r
revenue[1]
tickers[1]
```

If you want to __extract more than one item__ from the vector, you can indicate the positions within a vector and place it inside square brackets. Alternatively, if the positions follow a range of numbers, then you can simply create that sequence inside the square bracket. Here are two examples:
```r
revenue[c(2,6,10)]
tickers[1:5]
```

Operators can be directly applied to the vectors to perform operations on each element of the vector, as long as the operation is possible. 
```r
x <- c(1,2,3)
y <- c(4,5,6)
print(x+1)
print(y+2)
print(x*y)
print(y/x)
```

**Exercise**

1. Create a vector of numeric type that contains range of numbers from 1 to 30. Add a value of 1 to each item in this vector and store it in a new vector. Finally, determine the length of this new vector.

```{r}
# your code here
```

2. Create a vector of numeric type that contains 30 values of the number 3. Multiply the vector you created in Ex. #1 with this vector.

```{r}
# your code here
```


3. Run the following operation in r and inspect the result. 
```r
c(1,3) - c(2,4,6,8)
```


4. In your R console type `?c`. This will bring up the documentation for the function `c()`. Read the documentation to understand the function in detail.

```{r}
# your code here
```

### Matrices and Arrays
A matrix is a vector that has additional attributes called __dimensions__. Matrices have only two dimensions. Arrays are multidimensional matrices. Again, all elements in a matrix or array must be of same data type.  

Matrix and array can be created by using the `matrix()` and `array()` functions. Below is an example:
```r
amatrix <- matrix(1:6, nrow = 3, byrow=TRUE)
print(amatrix)
```
The above line of code creates a matrix containing numbers from 1 to 6, as indicated by `1:6`. It has three rows, as indicated by `nrow = 3` and the values 1 through 6 are filled row-wise, as indicated by `byrow = TRUE`. You may define the number of columns as well using `ncol=2`. However, if you don't R will try to figure it out based on the number of values you passed. 

When using the array function, the number of dimensions can be defined using `dim`. The example below creates a 3 dimensional array. It has two rows, three columns and four of those 2x3 matrices.
```r
my_array <- array(1:12, dim = c(2, 3, 4))
print(my_array)
```

It can be useful to give names to rows and columns when working with a matrix. 
```r
rownames(amatrix) = c('row1', 'row2', 'row3')
colnames(amatrix) = c('col1', 'col2')
print(amatrix)
```

Matrix operations can be performed by using the arithmetic operators on matrices. 
```r
amatrix + 10       # add scalar to each element
amatrix * 2        # multiply each element by scalar
amatrix + amatrix  # matrix addition
amatrix * amatrix  # element by element products
```

__Matrix multiplication__ can be performed using `%*%` operator as long as the operation is mathematically possible:
```r
bmatrix <- matrix(1:4, nrow=2)
amatrix%*%bmatrix
```

__Extracting elements__ from a matrix or array is similar to extracting elements from a vector. Additionally, since matrix and array have 2 or more dimension, each dimension is separated by a `,`.  
```r
print(bmatrix[1])
print(bmatrix[4])

print(bmatrix[,2])
print(bmatrix[2,2])

print(amatrix[2:3,1:2])
```

**Exercise**

1. Create a matrix from vector `c(1:4)` and store it as a. Create another matrix from vector `c(5:20)` and store is as b. 

```{r}
# your code here
```

2. Is it possible to perform matrix multiplication between a and b? If so, write the code below.

```{r}
# your code here
```

3. Create a 3-dimensional array from vector `c(1:12)`.

```{r}
# your code here
```

### Lists  
Lists are useful because they can store multiple data types inside of it. In fact we can even store matrices, arrays or list itself if we need to. Lists are ideal for storing non-rectangular data. `List()` function can be used to create list. 
```r
my_list <- list(aapl, amatrix, my_array)
print(my_list)
```
Each element of a list can be named during its constructions:
```r
my_list <- list(char = aapl, 
                mat = amatrix, 
                arr = my_array)
print(my_list)
```
Now we can use the `$` operator to access individual item in the list using its name:
```r
my_list$mat
```

**Exercise**

1. Create a list containing all data types and data structures we have learned so far.  

```{r}
# your code here
```

### Data frames
Data frame are the most commonly used data structure to store data. It is a powerful two-dimensional object comprising rows and columns. The difference between data frame and matrix is that data frame can store mixture of data types. It is similar to an excel file. An easy way to understand is that data frames are basically a bunch of vectors of the same length each with their own data types concatenated together. Below let's create a data frame from three vectors:
```r
revenue <- c(386.1, 286.2, 274.5, 182.5, 143, 119.5, 86, 82.6, 31.5, 21.9)
tickers <- c('AMZN', 'BRK.B', 'AAPL', 'GOOG', 'MSFT', 'JPM', 'FB', 'JNJ', 'TSLA', 'V')
company <- c('Amazon', 'Berkshire', 'Apple', 'Google', 'Microsoft', 'JP Morgan', 'Facebook', 'Johnson & Johnson', 'Tesla', 'Visa')

df <- data.frame(revenue_inbillions=revenue, stock_ticker=tickers, company_name=company)
print(df)

```
Notice that each column has a name and so do the rows. The name of the rows is simply the integer number that the row starts with. Various functions are available to explore the properties of data frame:
```r
dim(df)
str(df)
```
The `dim` function tells us that the given data frame is a 2-d data frame with 10 rows and 3 columns. The `str` function is an extremely useful function that provides a compact summary of the structure of the data frame object. It provides not only the dimension of the data frame but also the type of object each column is comprised of. It also prints out the first five elements of each column. 

Note: The `str` function can in fact be used with any R object. Try it out yourself and review the results. 

__Extracting elements__ from a data frame is similar to that of matrix. 
```r
df[,1]
df[1,]
df[5:10 , 1:2]
```

Data frames are called tidy data because each column have the same number of elements. Missing data in data frames are recorded as NA. You will be working with data frames a lot as real world data are mostly in tabular format and therefore, can be easily stored as a data frame in R. 

<!-- 
To import data into R, you can go to the "File" menu and click on "Import Dataset" option, where you will see further options to import various types of data. Or you may simply write a code to import the data. Many functions exist to import data in R. Each function is designed to import data stored differently. The data may be stored locally in your computer as a excel, csv or some other type of file. The data may also be found on the internet or obtained through other sources. 

Below is an example of importing data from a csv file using the function `read.csv` while providing the full path of the location of the data:
```r
PATH <- 'data/
df <- read.csv("https://s3.amazonaws.com/assets.datacamp.com/blog_assets/test.csv",
               header = TRUE)
```
-->

